\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-30 chapters/bci.tex
Started patients paper
Compiling

\todo{comments liesel paradigms}
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning}

@@

The most basic ability any independent human being needs, is interaction with
their environment.
In it's core essence, this is the ability to communicate your intentions,
emotions, frustrations or thoughts such that they might be acted on.
For an able-bodied person, physical interaction happens through body movements
while communication is usually done through speech and body language
Both require sending signals through the body's nervous system to control
muscles.
Not everyone has this capability.

The key problem for heavily paralyzed patients is that the mind wants to go
where the body can't follow.
Couldn't we therefore design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of
technology, like a robot arm or speech synthesizer, interacting with the world
on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also
termed brain-machine interface (BMI) when coupled to a physical actuator, like
a robot arm or an exoskeleton. The term \ac{bci} is usually preferred for
assistive technology and communication devices.}.
A \ac{bci} directly couples actions in the `real world' to the mental state of
the user.
%A \ac{bci} reads your mind.

Starting in the 1970's from methods to establish basic communication from
minimal brain activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities
by bypassing defective parts of the nervous or muscle system.
By now, classic \acp{bci} methods have matured outside the lab setting and can be
used as assistive communication technology by patients with impaired
speech~\cite{Wolpaw2018}.
Impressive cutting edge examples of experimental \acp{bci} include a brain-spine
interface allowing a paraplegic patient to walk again\cite{Lorach2023},
a \ac{bci} translating internal speech to a facial avatar speaking the imagined
text~\cite{Metzger2023}, and fast communication through decoding imagined
handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

\todo{put focus on communication BCI}

@@

Brain-computer interfacing is an entertaining subject with many futuristic and more established
applications.
Devices processing direct inputs from the central nervous system can be
useful for rehabilitation and medical diagnosis or treatment, but
have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability.
Furthermore, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.

Simply put, a \ac{bci} records the user's brain activity, extracts some
relevant output from this brain activity and couples this output to a function
of a device.\todo{refer to fig loop}
Optionally, the user can then observe the action of the device and adapt their
behavior or brain activity accordingly, closing the loop
\footnote{%
  \textcite{BCISociety2024} has recently formalized this into the
  following definition:
  \it``A brain-computer interface is a system that measures brain activity and
  converts it in (nearly) real-time into functionally useful outputs to replace,
  restore, enhance, supplement, and/or improve the natural outputs of the brain,
  thereby changing the ongoing interactions between the brain and its external or
  internal environments. It may additionally modify brain activity using targeted
  delivery of stimuli to create functionally useful inputs to the
  brain.''
}.

@@

The recorded brain waves form only one part of the interaction scheme presented
in Figure~\ref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
It would help greatly if we knew what we are looking for.
For this reason, a \ac{bci} operates using a defined \emph{paradigm}, the manner in
which interaction is performed.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
that the brain response evoked by attended stimuli is different of those
unattended,
and determine which one of a set of targets was attended.
In turn, this can be coupled to an action, like typing the inteded letter A or B
in Figure~\cite{fig:bci/loop}.
The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci/paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific
brain signatures related to our paradigm need to be identified within the recorded
activity.
\todo{preprocessing}
This is commonly referred to as \emph{feature extraction}.
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own brain activity.
On the other hand, the activity of interest often
originates from a specific region or network within the brain.
This activity might not easily be discerned from all other brain and
environmental activity.
Compare this to trying to listen to a single speaker at a noisy party where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out by
\emph{preprocessing}, but the problem of retrieving or \emph{decoding} only the
relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder.\todomvh{the decoder needs to take into
account the control facilities of the application (degrees of freedom, serial
/parallel nature, reactive/active...).}
Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
techniques.

Finally, the loop can be closed by coupling this to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.\todo{put focus on communication BCI}
On the other hand, the actions performed by the \ac{bci} can themselves alter
the user's cognitive state or brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as a hearing prosthesis.
More indirectly, modulation can be achieved by sensory stimulation.
This involves specific, paradigm-related sensory input, like haptic
feedback\todomvh{or tactile feedback:
S.N. Flesher et al., “A brain-computer
interface that evokes tactile sensations improves robotic arm control,”
Science, vol. 372, no. 6544, pp. 831-836, 2021. } for movement \ac{bci}, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural
pathways, can have a positive impact on \ac{bci} performance and gives rise to
rehabilitation applications.
\todo{is every section referred to?}

\todo{wrap-up}

\todo{check excercise session presentation}

\section{Recording technologies}
\label{sec:bci-recording}
\todo{make sure every part ends with a motivation for why we chose our
approach}

\todo{why is spatial resolution a good thing}
\todo{explain with each method what it actually is (e.g.\ electrodes placed
at...)}

The brain's activity can be recorded using various neuroimaging
technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
(fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
(fUS)~\cite{Zheng2023}\todomvh{non-invasive but not necessarily fUS...} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.
The technologies above all measure the hemodynamic signal, which only indirectly
reflects brain activity.
Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, the necessary
equipment is rather expensive and impractical.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/bci/recording_modalities.pdf}
}{%
	Different electrophysiology recording modalities and \\
	their respective position in a cross-section of the skull. \\
	There is a trade-off between invasiveness and  \\
	spatial resolution.
}{%
  fig:bci/recording
}
\todo{find better figure with permission to reuse and without unnecessary
modalities}

Specifically, invasive methods come with an increased \emph{spatial resolution},
meaning they can extract a signal from a specific brain region or even a set of
neurons of interest.
Generally, penetrating microelectrodes are considered to have the highest
spatial resolution.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular
potential of a small group of neurons, or even intracellular single neuron
action potential spikes.
They come in single-electrode form, or, more popularly, in microelectrode
arrays.
Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
NeuraLink implant~\cite{Musk2019}.\todomvh{+ Neuropixels 2}

@@

Some impressive recent breakthroughs in speech and motor \ac{bci} for communication
\todo{cite} have been realized using
invasive mictroelectrode arrays~\cite{Willett2021} and $\mu$\Ac{ecog}\cite{Metzger2023}.
Furthermore, recent advances in recording technology focus on improving implant
durability and resolution\todo{cite flexible}, and on balancing the invasiveness
tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered
through the bloodstream via a catheter, removing the need for open brain
surgery.
Nevertheless, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
potential communication \ac{bci} user groups consistently show that non-invasive
technology is preferred over implanted electrodes, unless the added value of invasive BCI is
sufficiently large~\cite{Kageyama2020}.\todo{due to what reasons}
This provides justification for focusing part of the current research effort on
non-invasive BCI.

@@

\section{\ac{bci} paradigms}
\label{sec:bci/paradigms}

@@

The \ac{bci} paradigms mentioned above can also be combined to gain performance.
Straightforward examples are activating or deactivating \ac{ssvep} stimulation
with a motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
along multiple dimensions at once.
\textcite{Han2023} recently used this strategy, combining frequency and phase
coding in \ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an
efficient \ac{bci} where one of 200 targets can be accurately selected with
only 800ms of stimulation.
\section{Decoders}
\label{sec:bci/decoding}

After preprocessing and extracting features, machine learning algorithms can perform \ac{bci}
decoding.
This is usually based on
some training data for which the performed tasks are known, by recognizing the
activity evoked by the task in recorded data.
The training data can be either obtained from the \ac{bci} user themselves or from
other users.
In the first case, the user would perform a short calibration session before
using the \ac{bci}, where they are instructed to perform a known task.
This calibration session can be eliminated by training the decoder on
preexisting labeled data from other sessions and users, but this is harder due
to the variability between subjects and sessions.\todo{make shorter, move part
to section}

\todo{regression/classification}
\todo{for communication, we want classification, focus on classification
further down the line}

@@

\textcite{Lotte2018, Xu2021} present a relatively recent overview of state-of-the art
decoder algorithms.
Classic linear methods, such as \ac{csp} feature extraction for
motor imagery~\cite{Park2017} and \ac{cca} for \ac{ssvep}~\cite{Nakanishi2017},
and \ac{lda} for \ac{erp} classification~\cite{Sosulski2022} still perform
relatively well, given some regularizing constraints or extensions.
Multi-linear techniques exploiting the tensor structure of neural signals are
also promising~\cite{Lotte2018}.
Riemannian Geometry~\cite{Barachant2014} is a popular and robust new strategy.
Furthermore, they lend themselves well to application like adaptive
learning~\cite{Benaroch2021} or transfer learning~\cite{Zanini2017}.
Riemannian Geometry classifiers are often considered the current
state-of-the-art.
Finally, deep learning~\cite{Bhuvaneshwari2021} is also sometimes considered,
albeit only when sufficiently large training datasets are available.
If decoders tailored to a specific user that keep the calibration session as
short as possible are of interest, not enough training data is available to
properly train a deep learning model.

@@

>>>>

\todo{comments liesel paradigms})
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}

@@

The most basic ability any independent human being needs, is interaction with
their environment.
In its core essence, this is the ability to communicate your intentions,
emotions, frustrations or thoughts such that they might be acted on.
For an able-bodied person, physical interaction happens through body movements
while communication is usually done through speech and body language
Both require sending signals through the body's nervous system to control
muscles.
Not everyone has this capability.

The key problem for severely paralyzed patients is that the mind wants to go
where the body can't follow.
Couldn't we therefore design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of
technology, like a robot arm or speech synthesizer, interacting with the world
on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also
termed brain-machine interface (BMI) when coupled to a physical actuator, like
a robot arm or an exoskeleton.
The term \ac{bci} is usually preferred for
assistive technology and communication devices.
},
a system that directly couples actions in the `real world' to the mental state of
the user.

Starting in the 1970's from methods to establish basic communication from
minimal brain activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities
by bypassing defective parts of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab
setting and can be used as assistive communication technology by patients with
impaired speech~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting edge examples of experimental use include a brain-spine
interface allowing a paraplegic patient to walk again\cite{Lorach2023},
a \ac{bci} translating internal speech to a facial avatar speaking the imagined
text~\cite{Metzger2023}, and fast communication through decoding imagined
handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

In general, devices processing direct inputs from the central nervous system
are useful for rehabilitation and medical diagnosis or treatment.
Additionally, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.\todo{cite}
Ultimatly, they have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability.
It is this \ac{bci} controlled communication is where our focus lies,
as it offers a transformative means for individuals with severe motor
impairments to regain functional independence to write and speak through
assistive technology.
This allows them to talk to their loved ones and caregivers, exercise hobbies
and makes many aspects of daily life more accessible.

@@

Simply put, a \ac{bci} records the user's brain activity, extracts some
relevant output from this brain activity and couples this output to a function
of a device as illustrated in Figure~\ref{fig:bci/loop}
Optionally, the user can then observe the action of the device and adapt their
behavior or brain activity accordingly, closing the loop\footnote{%
  \textcite{BCISociety2024} has recently formalized this into the
  following definition:
  \it``A brain-computer interface is a system that measures brain activity and
  converts it in (nearly) real-time into functionally useful outputs to replace,
  restore, enhance, supplement, and/or improve the natural outputs of the brain,
  thereby changing the ongoing interactions between the brain and its external or
  internal environments. It may additionally modify brain activity using targeted
  delivery of stimuli to create functionally useful inputs to the
  brain.''
}.

@@

The recorded brain waves form only one part of the interaction scheme presented
in Figure~\ref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
It would help greatly if we knew what we are looking for.
For this reason, a \ac{bci} operates using a defined \emph{paradigm}, the manner in
which interaction is performed.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
that the brain response evoked by attended stimuli is different of those
unattended,
and determine which one of a set of targets was attended.
In turn, this can be coupled to an action, like typing the inteded letter A or B
in Figure~\ref{fig:bci/loop}.
The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci/paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific
brain signatures related to our paradigm need to be identified within the recorded
activity.
This is commonly referred to as \emph{feature extraction}.
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own brain activity.
On the other hand, the activity of interest often
originates from a specific region or network within the brain.
This activity might not easily be discerned from all other brain and
environmental activity.
Compare this to trying to listen to a single speaker at a noisy party where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out by
\emph{preprocessing} (see Section~\ref{sec:bci/preprocessing}), but the problem of retrieving or \emph{decoding} only the
relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder.
Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
techniques.

Finally, the loop can be closed by coupling this to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.
For a communication system, this means coupling the decoded information to
e.g.\ a virtual keyboard or speech synthesizer.
On the other hand, the actions performed by the \ac{bci} can themselves alter
the user's cognitive state or brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as a hearing prosthesis.
More indirectly, modulation can be achieved by sensory stimulation.
This involves specific, paradigm-related sensory input, like tactile
feedback for movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural
pathways, can have a positive impact on \ac{bci} performance and gives rise to
rehabilitation applications.
\todo{is every section referred to?}

\todo{wrap-up}
\todo{make shorter}

\section{Recording technologies}
\label{sec:bci-recording}
\todo{why is spatial resolution a good thing}
\todo{explain with each method what it actually is (e.g.\ electrodes placed
at...)}
\todo{make sure every part ends with a short summary motivation for why we chose our
approach}

The brain's activity can be recorded using various neuroimaging
technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
(fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.
The technologies above all measure the hemodynamic signal, which only indirectly
reflects brain activity.
Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, the necessary
equipment is rather expensive and impractical.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/bci/recording_modalities.pdf}
}{%
	Different electrophysiology recording modalities and \\
	their respective position in a cross-section of the skull. \\
	There is a trade-off between invasiveness and  \\
	spatial resolution.
}{%
  fig:bci/recording
}

Specifically, invasive methods come with an increased \emph{spatial resolution},
meaning they can extract a signal from a specific brain region or even a set of
neurons of interest.
Generally, penetrating microelectrodes are considered to have the highest
spatial resolution.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular
potential of a small group of neurons, or even intracellular single neuron
action potential spikes.
They come in single-electrode form, or, more popularly, in microelectrode
arrays.
Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
NeuraLink implant~\cite{Musk2019} and IMEC's Neuropixels
2.0~\cite{Steinmetz2021}.

@@

Some impressive recent breakthroughs in speech and motor \ac{bci} for communication
have been realized using
invasive mictroelectrode arrays~\cite{Willett2021} and $\mu$\Ac{ecog}\cite{Metzger2023}.
Furthermore, recent advances in recording technology focus on improving implant
durability and resolution~\cite{Steinmetz2021}, and on balancing the invasiveness
tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered
through the bloodstream via a catheter, removing the need for open brain
surgery.
Nevertheless, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
potential communication \ac{bci} user groups consistently show that non-invasive
technology is preferred over implanted electrodes, unless the added value of invasive BCI is
sufficiently large~\cite{Kageyama2020}.\todo{due to what reasons}
This provides justification for focusing part of the current research effort on
non-invasive BCI.

@@

\section{\Ac{bci} paradigms}
\label{sec:bci/paradigms}

@@

The \ac{bci} paradigms mentioned above can also be combined to gain performance.
Straightforward examples are activating or deactivating \ac{ssvep} stimulation
with a motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
along multiple dimensions at once.
\textcite{Han2023} recently used this strategy, combining frequency and phase
coding in \ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an
efficient \ac{bci} where one of 200 targets can be accurately selected with
only 800ms of stimulation.

\section{Preprocessing}
\label{sec:bci/preprocessing}
\todo{preprocessing}
EEG preprocessing is a critical step in brain-computer interface (BCI) systems, as it
significantly improves the quality of the recorded data for subsequent analysis and
classification. Common preprocessing techniques in BCI systems include
rereferencing, band-pass and notch filtering, and \ac{ica}.

Rereferencing is an essential first step, where the potential of each EEG channel is
recalculated relative to a common reference point, such as the average of all
electrodes (common average reference) or a selected neutral set of electrodes.
This technique helps reduce noise common across all channels, improving signal clarity
and enhancing the accuracy of subsequent processing steps.

Band-pass filtering is then usually applied to isolate the frequency ranges of interest by
removing frequencies that are too high or too low to contain relevant neural signals.
Band-pass filters eliminate slow drifts and high-frequency noise, such as
muscle artifacts or environmental electrical interference, improving the
\ac{snr}.
Additionally, notch filtering can be  used to remove strong power line
interference\footnote{Typically at 50Hz in Europe}.
This interference can heavily contaminate EEG recordings and even leak through
when outside the passband of a previous filter.
Notch filtering efficiently attenuates this narrowband noise without affecting
the surrounding frequencies.

Finally, to address artifacts like eye blinks and saccades that can significantly
distort the EEG signal, \ac{ica} is
performed~\cite{Delorme2007}.
\Ac{ica} is a blind source separation technique that decomposes the EEG data into
statistically independent components. Components corresponding to eye movements
, blinks and other bodily or environmental artifacts can be identified, either
visually tased on their characteristic topographies and time courses, or
statistically, and then removed.
This leads to cleaner data for subsequent processing, especially in the case of
visual \ac{bci} paradigms.
Since eye movement plays a role in the paradigm, it can strongly be present in
the signal and correlate with the task, forming a confounding factor for
studying control based solely on brain activity.

Together, these preprocessing steps ensure that only the relevant neural signals are
passed on to the feature extraction and classification stages in BCI applications,
improving the accuracy and reliability of the system.

\section{Decoders}%
\label{sec:bci/decoding}

After preprocessing and extracting features, machine learning algorithms can perform \ac{bci}
decoding.
Machine learning decoders use some training data for which the performed tasks
are known, and learn to recognize the activity elicited by the task in recorded data.
The training data can be either obtained from the \ac{bci} user themselves or from
other users.
In the first case, the user would perform a short calibration session before
using the \ac{bci}, where they are instructed to perform a known task.
This calibration session can be eliminated by training the decoder on
preexisting labeled data from other sessions and users, but this is harder due
to the variability between subjects and sessions.

@@

Decoder choice is heavily dependent on constraints set by the paradigm and the
application.
The number of output degrees of features, the serial or parallel nature of
performed tasks, whether the paradigm is active or reactive, etc.\ all need to
be taken into account.
On of these choices is to solve a \emph{regression} or \emph{classification}
problem.
In regression, a continuous output variable is predicted for a new sample,
while classification attempts to sort it into one of a set of discrete classes.
Regression can be useful in \acp{bci} in applications like mapping imagined
movement to an external robot arm.
For communication \acp{bci}, however, conveying information in a symbolic
manner is often of interest.
Many applications resemble virtual keyboards or map a user's actions to
discrete words, predefined actions or letters for full control.
Hence, such problems often present themselves as classification tasks.

\textcite{Lotte2018, Xu2021} present a relatively recent overview of state-of-the art
classification algorithms for decoding.
Classic linear methods, such as \ac{csp} feature extraction for
motor imagery~\cite{Park2017} and \ac{cca} for \ac{ssvep}~\cite{Nakanishi2017},
and \ac{lda} for \ac{erp} classification~\cite{Sosulski2022} still perform
relatively well, given some regularizing constraints or extensions.
Multi-linear techniques exploiting the tensor structure of neural signals are
also promising~\cite{Lotte2018}.
Riemannian Geometry~\cite{Barachant2014} is a popular and robust new strategy.
Furthermore, they lend themselves well to application like adaptive
learning~\cite{Benaroch2021} or transfer learning~\cite{Zanini2017}.
Riemannian Geometry classifiers are often considered the current
state-of-the-art.
Finally, deep learning~\cite{Bhuvaneshwari2021} is also sometimes considered,
albeit only when sufficiently large training datasets are available.
If decoders tailored to a specific user that keep the calibration session as
short as possible are of interest, not enough training data is available to
properly train a deep learning model.

@@

\subsection{Preprocessing}
In addition to (ref) epoching
baseline correction, but not in decoding (cite)

### 2024-09-30 chapters/patients.tex
Compiling

\section{Introduction}

Communication BCIs have a target population consisting of patients in various
stages of paralysis or \ac{lis}, often suffering from oculomotor impairments.
This reduces their performance operating visual oddball \ac{bci}s (see
section\todo{ref section} for an overview), since they
are presumed unable to comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.\todo{cite some visual gaze-independent bci}
They might more naturally operate in covert \ac{vsa}, where the gaze and
\ac{vsa} do not coincide.
Several studies show that performance drops when not fixating the intended
target~\cite{Brunner2010, Frenzel2011,
Treder2010, RonAngevin2019, VanDenKerchove2024} \todo{[2, 1], our
study}, necessitating gaze-independent solutions.

@@

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table.tex}
  \caption{Presentation of included patients including their diagnosis and
  capabilities. (Trach.: patient underwent a tracheotomy, Cls.: classification according to~\textcite{Wolpaw2006}).
  }
  \label{tab:patients/patients}
\end{table}
\todo{Replace with classification according to 40. Kübler A, Birbaumer N.
Brain-computer interfaces and communication in paralysis: extinction of goal
directed thinking in completely paralysed patients? Clin Neurophysiol.
2008;119:2658-2666.}
\todo{include year of diagnosis}

@@

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table_eye.tex}
  \caption{Vision and eye motor impairment for included patients.
  Eye motor impairment was assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hasan2022} test.
  LogMAR: lower is better.}
  \label{tab:patients/eye}
\end{table}

>>>>

\section{Introduction}
\todo{rename patient to bci user and participant}

Communication BCIs have a target population consisting of patients in various
stages of paralysis or \ac{lis}, often suffering from oculomotor impairments.
This reduces their performance operating visual oddball \ac{bci}s (see
section\todo{ref section} for an overview), since they
are presumed unable to comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.\todo{cite some visual gaze-independent bci}
They might more naturally operate in covert \ac{vsa}, where the gaze and
\ac{vsa} do not coincide.
Several studies show that performance drops when not fixating the intended
target~\cite{Brunner2010, Frenzel2011, Treder2010, RonAngevin2019,
VanDenKerchove2024}, necessitating gaze-independent solutions.

@@

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table.tex}
  \caption[Presentation of included patients including their diagnosis and
  capabilities.]{Presentation of included patients including their diagnosis and
  capabilities.
  (Trach.: patient underwent a tracheotomy, Cls.: classification according
  to~\textcite{Wolpaw2006}).
  }
  \label{tab:patients/patients}
\end{table}
\todo{Replace with classification according to 40. Kübler A, Birbaumer N.
Brain-computer interfaces and communication in paralysis: extinction of goal
directed thinking in completely paralysed patients? Clin Neurophysiol.
2008;119:2658-2666.}
\todo{include year of diagnosis}

@@

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table_eye.tex}
  \caption[Vision and eye motor impairment for included patients.]{%
  Vision and eye motor impairment for included patients.
  Eye motor impairment was assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hassan2022} test.
  LogMAR: lower is better.}
  \label{tab:patients/eye}
\end{table}

### 2024-09-30 include/document.tex
Refactor structure
Comments François, smearing figure
Compiling

% ATTENTION: Don't build this, either build main_booklet.tex or
% main_manuscript.tex
%
% Preamble ====================================================================
\pdfminorversion=7
\usepackage{todonotes}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage[inline]{enumitem}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algpseudocode}

% Commands
\input{math.tex}
\input{acronyms.tex}
\newcommand{\includechapter}[2]{
  \chapter{#1}
  \begin{refsection}
  \input{#2}
    \printbibliography[heading=subbibliography]
  \end{refsection}
  \clearpage
}

\newcommand{\comment}[1]{
  \todo[color=yellow]{#1}
}
\newcommand*\elide{\textup{[\,\dots]}}

% Bibliography
\usepackage[backend=biber, doi=false, url=false, isbn=false, eprint=false]{biblatex}
\addbibresource{references.bib}
\addbibresource{publications.bib}

% Set bibliography font size
\renewcommand*{\bibfont}{\footnotesize}

\usepackage[inline]{enumitem}

@@

\part{Algorithms \& Decoders}
\includechapter{Kronecker-structured linear decoding}{chapters/stbf_struct.tex}

\chapter{Tensor discriminant analysis}
\todo{insert tensor paper}
\todo{discuss Tensorizations coping with latency jitter to make link}

\includechapter{Classfier-based Latency Estimation with Woody iterations}{chapters/wcble.tex}

@@

\backmatter%
\chapter{Funding \& acknowledgments}
\todo{my grant}\todo{marc's projects}
\chapter{Personal contribution}
\todo{Juliette, Alixe, Kristl Claeys, Edward Schietekatte, Marc, François,
Hakim, Marie,Etienne, Philip Van Damme  }
\chapter{Conflict of interest}
The authors declare no conflict of interest.
\include{chapters/cv.tex}
\end{document}

>>>>

@@

\part{Algorithms \& Decoders}
\includechapter{Kronecker-structured linear decoding}{chapters/stbf_struct.tex}
\includechapter{Block-term tensor discriminant analysis}{chapters/bttda.tex}
\todo{discuss Tensorizations coping with latency jitter to make link}
\includechapter{Latency estimation methods}{chapters/wcble.tex}

@@

\backmatter%
\chapter{Funding \& acknowledgments}
\todo{my grant}\todo{marc's projects}
\chapter{Personal contribution}
\todo{Juliette, Alixe, Kristl Claeys, Edward Schietekatte, Marc, François,
Hakim, Marie,Etienne, Philip Van Damme  }
\chapter{Conflict of interest}
The authors declare no conflict of interest.
\include{chapters/cv.tex}

\end{document}
